Decompositions of relative Chevalley groups

### 2015-10-06 bass-kolster.tex
Shortened statement of Corollary 4.3, title and abstract

The estimate in the case $\Phi=\rB_\ell$ can be obtained in a similar way. \end{proof}
The Bass---Kolster decomposition can be iterated if the stability assumption allows, with a slight improvement of \cref{cor:bass-kolster-count}.
\begin{cor}\label{cor:bass-kolster-iterated}
Let $\Phi$ be a root system, $R$ a commutative ring and $I$ its ideal. Then under a suitable stability assumption every element of $\G(\Phi,R,I)$ is a product of an element of $\G(\langle\pm\alpha_\ell\rangle,R,I)\cong\SL(2,R,I)$, at most $\ell-1$ elements of the form $z_{-\alpha}(*,*)$, $\alpha$ the maximal root of some terminal subsystem containing $\alpha_\ell$, and at most $N$ elementary root unipotents $x_\alpha(s)$, $s\in I$. The stability assumption and the number $N$ are as follows:
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right|+\ell; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right|+2\ell; \\
\Phi=\rB_\ell,\rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right|-\ell.
\end{array}\]
\end{cor}
\begin{proof}
The improved estimate for $\Phi=\rA_\ell$ or $\rC_\ell$ is obtained by using the fact that it suffices to make only two or three additions respectively to make the shorter column unimodular in the first step of the Bass---Kolster decomposition.
\end{proof}
\begin{figure}[htb]\label{fig:bass-kolster}
\[
\begin{tikzcd}[row sep=tiny]
\rD_4 \arrow[out=0,in=120]{rd}{\asr\leqslant3} & & & \\
\rA_4 \arrow{r}{\sr\leqslant4} & \rA_3 \arrow{r}{\sr\leqslant3} & \rA_2 \arrow{r}{\sr\leqslant2} & \rA_1 \\
\rC_4 \arrow{r}{\sr\leqslant7} & \rC_3 \arrow{r}{\sr\leqslant5} & \rC_2 \arrow[out=0,in=-120,swap]{ru}{\sr\leqslant3} \\
\rB_4 \arrow{r}{\asr\leqslant3} & \rB_3 \arrow[out=0,in=-120,swap]{ru}{\asr\leqslant2} & &
\end{tikzcd}
\]
\caption{Reductions provided by the Bass---Kolster decomposition}
\end{figure}

>>>>

The estimate in the case $\Phi=\rB_\ell$ can be obtained in a similar way. \end{proof}
By iterating Bass---Kolster decomposition one obtains the following result.
\begin{cor}\label{cor:bass-kolster-iterated}
Under the assumptions stated below every element of $\G(\Phi,R,I)$ can be factored as a product of one element of 
$\G(\langle\pm\alpha_\ell\rangle,R,I)\cong\SL(2,R,I)$ and at most $N$ elements of $\mathcal{Z}(\Sigma_\ell)$, where
\[\begin{array}{l@{\quad}l@{\quad}l}
\Phi=\rA_\ell, & \sr(I)\leqslant 2, & N=3\left|\Phi^+\right|+2\ell - 1; \\
\Phi=\rC_\ell, & \sr(I)\leqslant 3, & N=3\left|\Phi^+\right|+3\ell - 1; \\
\Phi=\rB_\ell,\rD_\ell, & \asr(I)\leqslant 2, & N=4\left|\Phi^+\right|-1.
\end{array}\]
\end{cor}
\begin{proof}
A direct consequence of~\cref{cor:bass-kolster-count}.
The improved estimate for $\Phi=\rA_\ell$ (resp. $\rC_\ell$) follows from the fact that it takes only two (resp. three) additions to shorten 
the unimodular column in the first step of the proof of~\cref{thm:BassKolster}.
\end{proof}
\begin{figure}[htb]\label{fig:bass-kolster}
\[
\begin{tikzcd}[row sep=tiny]
\rD_4 \arrow[out=0,in=120]{rd}{\asr\leqslant3} & & & \\
\rA_4 \arrow{r}{\sr\leqslant4} & \rA_3 \arrow{r}{\sr\leqslant3} & \rA_2 \arrow{r}{\sr\leqslant2} & \rA_1 \\
\rC_4 \arrow{r}{\sr\leqslant7} & \rC_3 \arrow{r}{\sr\leqslant5} & \rC_2 \arrow[out=0,in=-120,swap]{ru}{\sr\leqslant3} \\
\rB_4 \arrow{r}{\asr\leqslant3} & \rB_3 \arrow[out=0,in=-120,swap]{ru}{\asr\leqslant2} & &
\end{tikzcd}
\]
\caption{Reductions provided by the Bass---Kolster decomposition}
\end{figure}

### 2015-10-06 paper.tex
Shortened statement of Corollary 4.3, title and abstract

\title [Parabolic factorizations of congruence subgroups] {Parabolic factorizations of principal congruence subgroups in Chevalley groups and their applications}
\keywords {Chevalley groups, relative subgroups, stability for $\K_1$. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

\begin{abstract} We formulate and prove relative analogues (i.\,e. analogues for principal congruence subgroups) of Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of congruence subgroups with respect to elementary generators and subgroups of type $\SL_2$.
We also study submaximal factorizations of Chevalley groups of type $\rA_\ell$ and $\rD_\ell$.
\end{abstract}

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms of certain subsets of $\mathcal{Z}$.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
%TODO: Now check the coefficients of \rk(\Phi) in the expression below
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. 
Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\end{enumerate}
\end{thm}
The particular case $\Phi = \rA_\ell$, $R = I = \mathbb{Z}$ of the above theorem is similar in spirit to a recent theorem of U.~Hadad (see ~\cite[Theorem~1.6]{Ha12}).

>>>>

\title {Decompositions of relative Chevalley groups}
\keywords {Chevalley groups, principal congruence subgroups, stability for $\K_1$, bounded generation, subsystem factorizations, $\SL_2$-factorizations. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

\begin{abstract} 
We formulate and prove relative analogues (i.\,e. analogues for principal congruence subgroups) of Gauss, Bass---Kolster and Dennis---Vaserstein decompositions.
As an application we obtain upper estimates for the width of principal congruence subgroups with respect to several families of generators.
%elementary generators and subgroups of type $\SL_2$ we also study submaximal factorizations of Chevalley groups of type $\rA_\ell$ and $\rD_\ell$.
\end{abstract}

@@

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms of certain subsets of $\mathcal{Z}$.
\begin{thm}\label{thm:width} Let $\Phi$ be an irreducible root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z}(\Pi))\leqslant 3|\Phi^+|+2\rk(\Phi)-1$;
\item Let $p$ be a prime integer, set $R = \mathbb{Z}[\sfrac{1}{p}]$ and assume $\Phi$ is classical. Then under the assumption of the Generalized Riemann Hypothesis one has
%TODO: Now check the coefficients of \rk(\Phi) in the expression below
\begin{alignat*}{2}
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 3|\Phi^+| + 2\rk(\Phi) + 1 & \text{ if } \Phi=\rA_\ell,\rC_\ell, \\
& W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell^-))\leqslant 4|\Phi^+| + \rk(\Phi) + 1 & \text{ if } \Phi=\rB_\ell,\rD_\ell;
\end{alignat*}
\item Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. 
Assume that $\Phi$ is classical of rank $\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell^-))$ is finite.
\end{enumerate}
\end{thm}
Notice that a recent result of U.~Hadad and D.~W.~Morris \cite[Theorem~1.6]{Ha12} implies finiteness of $W(\E(\Phi, R, I), \mathcal{Z}(\Phi^+))$
for $R=\mathbb{Z}$ and $\Phi=\rA_\ell$.
